---
title: "Top Industries"
output: html_notebook
---

## Cleaning

```{r}
## Prep dataset
#Cleaning
# * Remove "Statewide" summary rows
# * Remove Territories (but leave DC)
# * X Make County name consistent
# * X Split COUNTY into County and State
# * X Make LFO (Name of industry) consistent
# * coerce factor variables to be factors (NOT DONE YET)

# Add a date type field for the year. Maybe 3/15/YEAR

# Deal with inflation (Use Consumer Price Index)
# * merge in inflation factors and convert to 2019 dollars
# * Annual Payroll adjusted for inflation
# * Quarter 1 payroll adjusted for inflation

# Add Average wage rates 
# * Average wage rate for all industries by county
# * Average wage rate for a given industry by county
# * Average wage rate for all industries by state
# * Average wage rate for a given industry by state


library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)

# Create a not in operator:
`%notin%` <- Negate(`%in%`)

# Read in Census Data
df <- read.csv("FinalProjectDataset.csv")

# Remove summary "Statewide" rows
df<-df %>% filter(substr(df$COUNTYSTATE,1,9)!="Statewide")
# Removes 4,683 rows that are statewide aggregate numbers


## Remove Territories
# * 60 = American Samoa
# * 66 = Guam
# * 69 = Commonwealth of the Northern Mariana Islands
# * 72 = Puerto Rico
# * 78 = United States Virgin Islands
 
## But Keep Washington DC
# * 11 = District of Columbia
 
excludeList <- list(60,66,69,72,78) # If we want to get rid of DC add 11 to this list
df<-df %>% filter(!STATENUM %in% excludeList)
# Removes 21,604 rows that are records on the territories

## Make County Name consistent

# add a field to df that concatenates state and county nums (STCTYNUM)
#df<-mutate(df,STCTYNUM=paste(str_pad(df$STATENUM,2,pad="0"),str_pad(df$COUNTYNUM,3,pad="0"),sep="_"))
df<-mutate(df,STCTYNUM=paste(str_pad(df$STATENUM,2,pad="0"),str_pad(df$COUNTYNUM,3,pad="0"),sep=""))

# create df of 2019 county names, state names, and STCTYNUM
countiesNames <- df %>% filter(YEAR == 2019) %>% 
  group_by(COUNTYSTATE) %>%
  summarise(
    COUNTY19 = unique(COUNTYSTATE),
    COUNTYNUM = unique(COUNTYNUM),
    STATENUM = unique(STATENUM)
  ) %>% 
  #mutate(STCTYNUM=paste(str_pad(STATENUM,2,pad="0"),str_pad(COUNTYNUM,3,pad="0"),sep="_")) %>% # Add concat of state and county nums (with underscore)
  mutate(STCTYNUM=paste(str_pad(STATENUM,2,pad="0"),str_pad(COUNTYNUM,3,pad="0"),sep="")) %>% # Add concat of state and county nums
  separate(COUNTY19,c("COUNTY","STATE"),sep=", ") %>% # split county and state names
  select(COUNTY,STATE,STCTYNUM) %>%
  arrange(STCTYNUM)

# Add to countiesNames the 14 that are missing from 2019
Missing2019 <- data.frame(COUNTY = c("Prince of Wales Outer Ketchikan Census Area",
                                     "Skagway-Hoonah-Angoon Census Area",
                                     "Kusilvak Census Area", 
                                     "Wrangell-Petersburg Census Area",
                                     "Kalawao County", "Petroleum County",
                                     "Banner County", "McPherson County",
                                     "Esmeralda County", "Oglala Lakota County",
                                     "Borden County", "King County", 
                                     "Loving County", "Bedford City"),
                          STATE = c("Alaska", "Alaska", "Alaska", "Alaska", 
                                    "Hawaii", "Montana", "Nebraska", "Nebraska",
                                    "Nevada", "South Dakota", "Texas", "Texas",
                                    "Texas", "Virginia"),
#                          STCTYNUM = c("02_201", "02_232", "02_270", "02_280",
#                                       "15_005", "30_069", "31_007", "31_117",
#                                       "32_009", "46_113", "48_033", "48_269", 
#                                       "48_301","51_515"))
                          STCTYNUM = c("02201", "02232", "02270", "02280",
                                       "15005", "30069", "31007", "31117",
                                       "32009", "46113", "48033", "48269", 
                                       "48301","51515"))

countiesNames <- rbind(countiesNames, Missing2019)

# Add COUNTY and STATE to df
#df<-merge(df,counties2019, by="STCTYNUM", all=TRUE) # use this one to debug the missing records
df <- merge(df, countiesNames, by="STCTYNUM")

## Make LFO consistent
industries <- df %>% filter(YEAR == 2019) %>% 
  group_by(NAIC) %>%
  summarise(
    INDUSTRY = unique(LFO),
  ) %>% 
  arrange(INDUSTRY)

# Need to trim the whitespace off NAIC.
df$NAIC <- str_trim(df$NAIC, side="both")

#df<-merge(df, industries, by="NAIC")
df<-merge(df, industries, by="NAIC",all=TRUE)

## Add a date type field for the year. Maybe 3/15/YEAR

df$DATE <- mdy(paste("03", "15", df$YEAR, sep="/"))

##Deal with inflation (Use Consumer Price Index)
# * merge in inflation factors and convert to 2019 dollars
# * Annual Payroll adjusted for inflation
# * Quarter 1 payroll adjusted for inflation

# Add inflation multiplier to convert to 2019 dollars to df
CPI <- read.csv("CPI2019.csv")
CPI<-CPI[,c("Year","CPI_2019")] # removed Annual column before merge
df<-merge(df, CPI, by.x="YEAR", by.y="Year")

# Add fields for Annual Payroll and Quarterly Payroll in 2019 dollars
df$PAYANN_19DOL = df$PAYANN * df$CPI_2019
df$PAYQTR1_19DOL = df$PAYQTR1 * df$CPI_2019

## Average wage rates by year
# * Average wage rate for a given industry by county
# * Average wage rate by county, regardless of industry
# * Average wage rate for each state, regardless of industry
# * Average wage rate for an industry by state
 
# Round these new fields to 2 decimal places

# Average wage rate for counties, regardless of industry
df<-df %>% group_by(STCTYNUM) %>%
  mutate(CTYPAYRATE = round(sum(PAYANN)/sum(EMP), digits=2)) %>%  ## not adj for inflation
  mutate(CTYPAYRATE_19DOL = round(sum(PAYANN_19DOL)/sum(EMP), digits=2)) %>% ## adjusted to 2019 dollars
  ungroup()

#Average wage rate by Year-County-Industry, add to df
df$CTYPAYRATE_NAIC = round(df$PAYANN/df$EMP, digits=2)  ## not adj for inflation
df$CTYPAYRATE_NAIC_19DOL = round(df$PAYANN_19DOL/df$EMP, digits=2) ## adjusted to 2019 dollars

# Average wage rate for each state, regardless of industry
df <- df %>% group_by(STATENUM) %>%
  mutate(STATEPAYRATE = round(sum(PAYANN)/sum(EMP), digits=2))  %>%  ## not adj for inflation
  mutate(STATEPAYRATE_19DOL = round(sum(PAYANN_19DOL)/sum(EMP), digits=2)) %>% ## adjusted to 2019 dollars
  ungroup()

# Average wage rate for an industry by state
df<-df %>% group_by(STATENUM,NAIC) %>%
  mutate(STATEPAYRATE_NAIC = round(sum(PAYANN)/sum(EMP), digits=2)) %>% ## not adj for inflation
  mutate(STATEPAYRATE_NAIC_19DOL = round(sum(PAYANN_19DOL)/sum(EMP), digits=2)) %>% ## adjusted to 2019 dollars
  ungroup()

##Coerce to factors
# * NAIC
# * STCTYNUM
# * STATENUM
# * COUNTYNUM
# * COUNTY
# * STATE

df$NAIC <- as.factor(df$NAIC)
# Not factoring these because usmap wants them as characters
#df$STCTYNUM <- as.factor(df$STCTYNUM)
df$STATENUM <- as.factor(df$STATENUM)
df$COUNTYNUM <- as.factor(df$COUNTYNUM)
df$COUNTY <- as.factor(df$COUNTY)
df$STATE <- as.factor(df$STATE)

# for usmap
df$FIPS_ST <-str_sub(df$STCTYNUM,1,2)
df$FIPS_CTY <-as.character(df$STCTYNUM)

# Remove fields no longer needed
df<- df %>% select(-c("COUNTYSTATE","LFO"))

str(df)

``` 

# Get Rid of industry 99

```{r}
df<-df%>%filter(NAIC!=99)
```


################################################################################


# Libraries for plotting



```{r}
library(ggplot2)
library(ggrepel)
library(usmap)
# https://rdrr.io/cran/usmap/f/vignettes/mapping.Rmd
# https://rdrr.io/cran/usmap/f/vignettes/advanced-mapping.Rmd

library(scales)
#?scales::trans_new
```

# Setup Color pallate for industries

```{r}
# Setup plotting colors by NAIC code

industries<-industries %>% arrange(INDUSTRY)

plotColor<-c(
"Accommodation and food services"="red",
"Administrative and support and waste management and remediation services"="blue",
"Agriculture, forestry, fishing and hunting"="pink",
"Arts, entertainment, and recreation"="MediumVioletRed",
"Construction"="Salmon",
"Educational services"="BlueViolet",
"Finance and insurance"="YellowGreen",
"Health care and social assistance"="SeaGreen",
"Industries not classified"="Tan",
"Information"="maroon",
"Management of companies and enterprises"="limegreen",
"Manufacturing"="DarkOliveGreen",
"Mining, quarrying, and oil and gas extraction"="springgreen",
"Other services (except public administration)"="orange",
"Professional, scientific, and technical services"="turquoise",
"Real estate and rental and leasing"="SteelBlue",
"Retail trade"="khaki",
"Transportation and warehousing"="orchid",
"Utilities"="DarkBlue",
"Wholesale trade"="RosyBrown"
)
industries<- cbind(industries, plotColor)

```

## usmap Mapping
You'll need to install the package: usmap

```{r}
## SKIP ##

#states_df <- usmap::us_map()
#counties_df <- usmap::us_map(regions = "counties")

# usmap's built in map plotting
plot_usmap(regions = "counties") + 
  labs(title = "US Counties",
       subtitle = "This is a blank map of the counties of the United States.") + 
  theme(panel.background = element_rect(color = "black", fill = "lightblue"))

```

# US CHOROPLETHS

## Top Annual Payroll by Industry by state by year

```{r}

# Top Annual Payroll by Industry by state by year
#Inflation Adj

thisYear = 2019 # Just change this line to customize for year

ToPlot <- df %>% filter(YEAR==thisYear) %>%
  group_by(FIPS_ST, INDUSTRY) %>%
  summarise(TotalPayAnn = sum(PAYANN_19DOL)) %>%
  filter(TotalPayAnn == max(TotalPayAnn)) %>% 
  merge(industries)
ToPlot<-rename(ToPlot, "fips"="FIPS_ST")

cols<-ToPlot%>%
  group_by(INDUSTRY)%>%summarize(pc=unique(plotColor))

plot_usmap(data = ToPlot, values="INDUSTRY") + 
  scale_fill_manual(values=cols$pc, name="Industries", labels=cols$INDUSTRY) +
  labs(title = paste(as.character(thisYear),"Top Industry by Annual Payroll in Each State",sep=" ")) + 
  theme(legend.position = "right")

```

# Top Number of Employees by Industry by state by year

```{r}

thisYear = 2019 # Just change this line to customize for year

ToPlot <- df %>% filter(YEAR==thisYear) %>%
  group_by(FIPS_ST, INDUSTRY) %>%
  summarise(TotalEmp = sum(EMP)) %>%
  filter(TotalEmp == max(TotalEmp)) %>% 
  merge(industries)
ToPlot<-rename(ToPlot, "fips"="FIPS_ST")

cols<-ToPlot%>%
  group_by(INDUSTRY)%>%summarize(pc=unique(plotColor))

plot_usmap(data = ToPlot, values="INDUSTRY") + 
  scale_fill_manual(values=cols$pc, name="Industries", labels=cols$INDUSTRY) +
  labs(title = paste(as.character(thisYear),"Top Industry by Number of Employees in Each State",sep=" ")) + 
  theme(legend.position = "right")

```
# Top Number of Establishments by Industry by state by year

```{r}

thisYear = 2019 # Just change this line to customize for year

ToPlot <- df %>% filter(YEAR==thisYear) %>%
  group_by(FIPS_ST, INDUSTRY) %>%
  summarise(TotalEstab = sum(ESTAB)) %>%
  filter(TotalEstab == max(TotalEstab)) %>% 
  merge(industries)
ToPlot<-rename(ToPlot, "fips"="FIPS_ST")

cols<-ToPlot%>%
  group_by(INDUSTRY)%>%summarize(pc=unique(plotColor))

plot_usmap(data = ToPlot, values="INDUSTRY") + 
  scale_fill_manual(values=cols$pc, name="Industries", labels=cols$INDUSTRY) +
  labs(title = paste(as.character(thisYear),"Top Industry by Number of Establishments in Each State",sep=" ")) + 
  theme(legend.position = "right")

```

# Highest Pay rate by Industry by state by year

```{r}

thisYear = 2019 # Just change this line to customize for year

ToPlot <- df %>% filter(YEAR==thisYear) %>%
  group_by(FIPS_ST, INDUSTRY) %>%
  summarise(PayRate = sum(PAYANN_19DOL)/sum(EMP)) %>%
  filter(PayRate == max(PayRate)) %>% 
  merge(industries)
ToPlot<-rename(ToPlot, "fips"="FIPS_ST")

cols<-ToPlot%>%
  group_by(INDUSTRY)%>%summarize(pc=unique(plotColor))

plot_usmap(data = ToPlot, values="INDUSTRY") + 
  scale_fill_manual(values=cols$pc, name="Industries", labels=cols$INDUSTRY) +
  labs(title = paste(as.character(thisYear),"Highest Pay rate by Industry in Each State",sep=" ")) + 
  theme(legend.position = "right")

```

# STATE CHOROPLETHS

# Top Annual Payroll by Industry in State by year

```{r}
thisState = "Iowa" # Just change this line to customize for state
thisYear = 2019 # Just change this line to customize for year

ToPlot <- df %>% filter(YEAR==thisYear, STATE==thisState) %>%
  group_by(FIPS_CTY, INDUSTRY) %>%
  summarise(TotalPayAnn = sum(PAYANN_19DOL)) %>%
  filter(TotalPayAnn == max(TotalPayAnn)) %>% 
  merge(industries)
ToPlot<-rename(ToPlot, "fips"="FIPS_CTY")

cols<-ToPlot%>%
  group_by(INDUSTRY)%>%summarize(pc=unique(plotColor))

plot_usmap(data = ToPlot, values="INDUSTRY", include=c("Iowa")) + 
  scale_fill_manual(values=cols$pc, name="Industries", labels=cols$INDUSTRY) +
  labs(title = paste(as.character(thisYear),"Top Industry by Annual Payroll in", thisState,sep=" ")) + 
  theme(legend.position = "right")

```
# Highest Pay rate by Industry in Each County by year

```{r}
thisState = "Iowa" # Just change this line to customize for state
thisYear = 2019 # Just change this line to customize for year

ToPlot <- df %>% filter(YEAR==thisYear, STATE==thisState) %>%
  group_by(FIPS_CTY, INDUSTRY) %>%
  summarise(PayRate = sum(PAYANN_19DOL)/sum(EMP)) %>%
  filter(PayRate == max(PayRate)) %>% 
  merge(industries)
ToPlot<-rename(ToPlot, "fips"="FIPS_CTY")

cols<-ToPlot%>%
  group_by(INDUSTRY)%>%summarize(pc=unique(plotColor))

plot_usmap(data = ToPlot, values="INDUSTRY", include=c("Iowa")) + 
  scale_fill_manual(values=cols$pc, name="Industries", labels=cols$INDUSTRY) +
  labs(title = paste(as.character(thisYear),"Highest Pay rate by Industry in", thisState,sep=" ")) + 
  theme(legend.position = "right")

```

# Top Number of Establishments by Industry in a State by year

```{r}
thisState = "Iowa" # Just change this line to customize for state
thisYear = 2019 # Just change this line to customize for year

ToPlot <- df %>% filter(YEAR==thisYear, STATE==thisState) %>%
  group_by(FIPS_CTY, INDUSTRY) %>%
  summarise(TotalEstab = sum(ESTAB)) %>%
  filter(TotalEstab == max(TotalEstab)) %>% 
  merge(industries)
ToPlot<-rename(ToPlot, "fips"="FIPS_CTY")

cols<-ToPlot%>%
  group_by(INDUSTRY)%>%summarize(pc=unique(plotColor))

plot_usmap(data = ToPlot, values="INDUSTRY", include=c("Iowa")) + 
  scale_fill_manual(values=cols$pc, name="Industries", labels=cols$INDUSTRY) +
  labs(title = paste(as.character(thisYear),"Top Industry by Number of Establishments in", thisState,sep=" ")) + 
  theme(legend.position = "right")

```

# Top Number of Establishments by Industry in a State by year
Staci

```{r}
thisState = "Iowa" # Just change this line to customize for state
thisYear = 2019 # Just change this line to customize for year

ToPlot <- df %>% filter(YEAR==thisYear, STATE==thisState) %>%
  group_by(FIPS_CTY, INDUSTRY) %>%
  summarise(TotalEstab = sum(ESTAB)) %>%
  filter(TotalEstab == max(TotalEstab)) %>% 
  merge(industries)
ToPlot<-rename(ToPlot, "fips"="FIPS_CTY")

cols<-ToPlot%>%
  group_by(INDUSTRY)%>%summarize(pc=unique(plotColor))

plot_usmap(data = ToPlot, values="INDUSTRY", include=c("Iowa")) + 
  scale_fill_manual(values=cols$pc, name="Industries", labels=cols$INDUSTRY) +
  labs(title = paste(as.character(thisYear),"Top Industry by Number of Establishments in", thisState,sep=" ")) + 
  theme(legend.position = "right")

```

# Line graphs of Annual Payroll by Industry
## For best results run the plot in the console window or an r script to be able to export or zoom into the plot
Export at 1200 x 800 to fit 2 plots per page

## Entire US - Annual Payroll by Industry
```{r}

# All States
# For best results run the plot in the console window or an r script to be able to export or zoom into the plot
ToPlot <- df %>%
  group_by(YEAR, NAIC) %>%
  summarise(TotalPayAnn = sum(PAYANN_19DOL)) %>% 
  merge(industries)

cols<-ToPlot%>%group_by(NAIC,INDUSTRY)%>%summarize(pc=unique(plotColor))

ToPlot %>% 
  mutate(label = if_else(YEAR == 2019, as.character(NAIC), NA_character_)) %>%
  ggplot(aes(x=YEAR, y=TotalPayAnn, color=NAIC)) +
  scale_colour_manual(values=cols$pc, name="Industries", labels=c(paste(cols$NAIC,cols$INDUSTRY,sep=": "))) +
  geom_line(size=1.25, alpha=.6) +
  scale_x_continuous(expand=c(0,0,0,2)) +
  scale_y_continuous(labels=dollar) +
  geom_label_repel(aes(label = label), nudge_x=2, na.rm=TRUE) +
  labs(title="United States: Total Annual Payroll by Industry (in 2019 dollars)", x="Year", y="Total Annual Payroll")

```

## Entire US - Number Employed by Industry
```{r}

# All States
# For best results run the plot in the console window or an r script to be able to export or zoom into the plot
ToPlot <- df %>%
  group_by(YEAR, NAIC) %>%
  summarise(TotalEmp = sum(EMP)) %>% 
  merge(industries)

cols<-ToPlot%>%group_by(NAIC,INDUSTRY)%>%summarize(pc=unique(plotColor))

ToPlot %>% 
  mutate(label = if_else(YEAR == 2019, as.character(NAIC), NA_character_)) %>%
  ggplot(aes(x=YEAR, y=TotalEmp, color=NAIC))+
  scale_colour_manual(values=cols$pc, name="Industries", labels=c(paste(cols$NAIC,cols$INDUSTRY,sep=": "))) +
  geom_line(size=1.25, alpha=.6) +
  scale_x_continuous(expand=c(0,0,0,2)) +
  scale_y_continuous(labels=comma) +
  geom_label_repel(aes(label = label), nudge_x=2, na.rm=TRUE) +
  labs(title="United States: Total Number Employed by Industry", x="Year", y="Employees")

```

## Entire US - Number of Establishments by Industry

```{r}

# All States
# For best results run the plot in the console window or an r script to be able to export or zoom into the plot
ToPlot <- df %>%
  group_by(YEAR, NAIC) %>%
  summarise(TotalEstab = sum(ESTAB)) %>% 
  merge(industries)

cols<-ToPlot%>%group_by(NAIC,INDUSTRY)%>%summarize(pc=unique(plotColor))

ToPlot %>% 
  mutate(label = if_else(YEAR == 2019, as.character(NAIC), NA_character_)) %>%
  ggplot(aes(x=YEAR, y=TotalEstab, color=NAIC))+
  scale_colour_manual(values=cols$pc, name="Industries", labels=c(paste(cols$NAIC,cols$INDUSTRY,sep=": "))) +
  geom_line(size=1.25, alpha=.6) +
  scale_x_continuous(expand=c(0,0,0,2)) +
  scale_y_continuous(labels=comma) +
  geom_label_repel(aes(label = label), nudge_x=2, na.rm=TRUE) +
  labs(title="United States: Total Number Establishments by Industry", x="Year", y="Establishments")

```

# Entire US: Average Pay Rates

```{r}

# All States
ToPlot <- df %>%
  group_by(YEAR, NAIC) %>%
  summarise(PayRate = sum(PAYANN_19DOL)/sum(EMP)) %>% 
  merge(industries)

cols<-ToPlot%>%group_by(NAIC,INDUSTRY)%>%summarize(pc=unique(plotColor))

ToPlot %>% 
  mutate(label = if_else(YEAR == 2019, as.character(NAIC), NA_character_)) %>%
  ggplot(aes(x=YEAR, y=PayRate, color=NAIC))+
  scale_colour_manual(values=cols$pc, name="Industries", labels=c(paste(cols$NAIC,cols$INDUSTRY,sep=": "))) +
  geom_line(size=1.25, alpha=.6) +
  scale_x_continuous(expand=c(0,0,0,2)) +
  scale_y_continuous(labels=dollar) +
  geom_label_repel(aes(label = label), nudge_x=2, na.rm=TRUE) +
  labs(title="United States: Average Pay Rate by Industry (in 2019 dollars)", x="Year", y="Avg. Pay Rate")

```
# Entire US: Average Employees per Establishment

```{r}

# All States
ToPlot <- df %>%
  group_by(YEAR, NAIC) %>%
  summarise(EERate = sum(EMP)/sum(ESTAB)) %>% 
  merge(industries)

cols<-ToPlot%>%group_by(NAIC,INDUSTRY)%>%summarize(pc=unique(plotColor))

ToPlot %>% 
  mutate(label = if_else(YEAR == 2019, as.character(NAIC), NA_character_)) %>%
  ggplot(aes(x=YEAR, y=EERate, color=NAIC))+
  scale_colour_manual(values=cols$pc, name="Industries", labels=c(paste(cols$NAIC,cols$INDUSTRY,sep=": "))) +
  geom_line(size=1.25, alpha=.6) +
  scale_x_continuous(expand=c(0,0,0,2)) +
  scale_y_continuous(labels=comma) +
  geom_label_repel(aes(label = label), nudge_x=2, na.rm=TRUE) +
  labs(title="United States: Average Employees per Establishment by Industry", x="Year", y="Employees per Establishment")

```

# State Line plots

## One state - Annual Payroll by Industry (in 2019 dollars)

```{r}

thisState ="Iowa"  #Only need to change this line to customize to a state

ToPlot <- df %>% filter(STATE==thisState) %>%
  group_by(YEAR, NAIC) %>%
  summarise(TotalPayAnn = sum(PAYANN_19DOL)) %>% 
  merge(industries)

cols<-ToPlot%>%group_by(NAIC,INDUSTRY)%>%summarize(pc=unique(plotColor))

ToPlot %>% 
  mutate(label = if_else(YEAR == 2019, as.character(NAIC), NA_character_)) %>%
  ggplot(aes(x=YEAR, y=TotalPayAnn, color=NAIC)) +
  scale_colour_manual(values=cols$pc, name="Industries", labels=c(paste(cols$NAIC,cols$INDUSTRY,sep=": "))) +
  geom_line(size=1.25, alpha=.6) +
  scale_x_continuous(expand=c(0,0,0,2)) +
  scale_y_continuous(labels=dollar) +
  geom_label_repel(aes(label = label), nudge_x=2, na.rm=TRUE) +
  labs(title=paste(thisState,"Total Annual Payroll by Industry (in 2019 dollars)",sep=": "), x="Year", y="Total Annual Payroll")

```

## One State - Number Employed by Industry
```{r}

thisState ="Iowa"  #Only need to change this line to customize to a state

ToPlot <- df %>% filter(STATE==thisState) %>%
  group_by(YEAR, NAIC) %>%
  summarise(TotalEmp = sum(EMP)) %>% 
  merge(industries)

cols<-ToPlot%>%group_by(NAIC,INDUSTRY)%>%summarize(pc=unique(plotColor))

ToPlot %>% 
  mutate(label = if_else(YEAR == 2019, as.character(NAIC), NA_character_)) %>%
  ggplot(aes(x=YEAR, y=TotalEmp, color=NAIC))+
  scale_colour_manual(values=cols$pc, name="Industries", labels=c(paste(cols$NAIC,cols$INDUSTRY,sep=": "))) +
  geom_line(size=1.25, alpha=.6) +
  scale_x_continuous(expand=c(0,0,0,2)) +
  scale_y_continuous(labels=comma) +
  geom_label_repel(aes(label = label), nudge_x=2, na.rm=TRUE) +
  labs(title=paste(thisState,"Total Number Employed by Industry",sep=": "), x="Year", y="Employees")

```


## One State - Number of Establishments by Industry

```{r}

thisState ="Iowa"  #Only need to change this line to customize to a state

ToPlot <- df %>% filter(STATE==thisState) %>%
  group_by(YEAR, NAIC) %>%
  summarise(TotalEstab = sum(ESTAB)) %>% 
  merge(industries)

cols<-ToPlot%>%group_by(NAIC,INDUSTRY)%>%summarize(pc=unique(plotColor))

ToPlot %>% 
  mutate(label = if_else(YEAR == 2019, as.character(NAIC), NA_character_)) %>%
  ggplot(aes(x=YEAR, y=TotalEstab, color=NAIC))+
  scale_colour_manual(values=cols$pc, name="Industries", labels=c(paste(cols$NAIC,cols$INDUSTRY,sep=": "))) +
  geom_line(size=1.25, alpha=.6) +
  scale_x_continuous(expand=c(0,0,0,2)) +
  scale_y_continuous(labels=comma) +
  geom_label_repel(aes(label = label), nudge_x=2, na.rm=TRUE) +
  labs(title=paste(thisState,"Total Number Establishments by Industry",sep=": "), x="Year", y="Establishments")

```


# One State: Average Pay Rates

```{r}

thisState ="Iowa"  #Only need to change this line to customize to a state

ToPlot <- df %>% filter(STATE==thisState) %>%
  group_by(YEAR, NAIC) %>%
  summarise(PayRate = sum(PAYANN_19DOL)/sum(EMP)) %>% 
  merge(industries)

cols<-ToPlot%>%group_by(NAIC,INDUSTRY)%>%summarize(pc=unique(plotColor))

ToPlot %>% 
  mutate(label = if_else(YEAR == 2019, as.character(NAIC), NA_character_)) %>%
  ggplot(aes(x=YEAR, y=PayRate, color=NAIC))+
  scale_colour_manual(values=cols$pc, name="Industries", labels=c(paste(cols$NAIC,cols$INDUSTRY,sep=": "))) +
  geom_line(size=1.25, alpha=.6) +
  scale_x_continuous(expand=c(0,0,0,2)) +
  scale_y_continuous(labels=dollar) +
  geom_label_repel(aes(label = label), nudge_x=2, na.rm=TRUE) +
  labs(title=paste(thisState,"Average Pay Rate by Industry (in 2019 dollars)",sep=": "), x="Year", y="Avg. Pay Rate")

```
